\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-10 doc/incertae_sedis.tex
pruned
up to last couple of pages

Note that this example also presents a challenge for iterative supertree building and refinement
    of the taxonomy without retention of additional information.
Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
    rooted split: $W\, V1 \mid \bullet\,V2$.
If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
While the topology for the final tree in either order of analyses would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}b}, the decision about whether or not to name $V$ would differ.
Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
As expected, in the encoding of the revised taxonomy, the definition of $WX$ has been altered to reflect the
    fact $T_1$ placed $V$ inside $WX$.
Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.

@@

\begin{figure}
\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/placement1/tax\lyxdot tre}

}\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/placement1/input1\lyxdot tre}

}\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/placement1/input2\lyxdot tre}

}

\caption{\label{fig:placement-example1}Example.
    An incertae sedis clade (ott6,ott7)
    is placed in different subtrees by input trees $\tau_{1}$ and $\tau_{2}$.
    In $\tau_{1}$, two nodes that correspond to the taxonomy their ingroup
    extended to include (ott6,ott7), and the branches leading to these
    nodes have been colored blue.
    The dashed blue edge leads to a node
    that is a newly-introduced degree-2 node which does not correspond
    to any taxonomy node.
    In $\tau_{2}$, only one node that corresponds
    to a taxonomy node needs to have its ingroup extended.
    The placement
    of (ott6,ott7) into ott8 toward ott1 by $\tau_{1}$ conflicts with
    the placement of (ott6,ott7) into ott9 toward ott3 by $\tau_{2},$}
\end{figure}

@@

@@

We must also handle conflicting placements of \emph{incertae sedis}
taxa by different input trees.
Thus, if one input tree places the
\emph{incertae sedis} taxon $X$ in $((X)B)A$ and another places $X$ in
$((X)C)A$ then we must mark both edges $B$ and $C$ as contested edges,
even if these edges would \emph{not} be contested were taxon $X$ to be
removed.
This results in a new way to contest edges that involves the
interaction of two input trees, and not just the interaction of each
input tree with the taxonomy.

We choose to solve these problems by merging any subproblems that an
\emph{incertae sedis} taxon might be placed in.
The simplest way to
achieve this is simply to regard any taxon that has an \emph{incertae
sedis }taxon placed within it as contested.
This results in marking
both $B$ and $C$ as contested edges in the example above.
In fact,
this is the current behaviour of the non-\emph{incertae-sedis} aware
subproblem decomposer.
One downside of this approach is that, if we
have $((X)B)A)$ in one input tree, and $X$ is mentioned nowhere else,
then by marking $B$ as contested, we are creating unnecessarily large
subproblems.
We could instead placed $X$ in $B$ and avoid contesting
the edge $B$.
However, this approach is more complex and does not seem
necessary in practice.

 For example, in Figure
\ref{fig:placement-example1}, input tree $\tau_{1}$ contests ott9 and
input tree $\tau_{2}$ contests ott9 and ott8.
Thus ott1, ott2, ott3,
ott6, and ott7 end up in the same sub-problem.

@@

\subsubsection{Reading incertae sedis information}

Currently, we read the \emph{incertae sedis} information as a list of
OTT ids for \emph{incertae sedis} taxa.
This does not require adding
further annotations to the node names.
Only taxonomy nodes can be
\emph{incertae sedis} at the moment, and only the taxonomy tree for
the subproblem contains OTT ids for internal nodes.
Therefore we
handle \emph{incertae sedis} information by constructing modified
split sets for the lowest-ranked tree when the list of \emph{incertae
sedis} nodes is not empty.

\subsubsection{Exclude sets modified by \emph{incertae sedis} marks}

Equation (\ref{eq:exsetformone}) leads to the following
algorithm to compute the exclude set for all nodes in a tree.
\begin{enumerate} \item Set the exclude set of the root node to be
empty \item For each \emph{node} (except the root) in preorder
\begin{itemize} \item combine the \emph{exclude} set of the parent
node with the \emph{include} set of non-\emph{incertae-sedis}
siblings.
\item store this set in a hash, with key \emph{node}
\end{itemize} \end{enumerate} This algorithm is currently implemented
in \emph{otc-solve- subproblem}.
We store the sets as \emph{std::set}.

>>>>

Note that this example also presents a challenge for iterative supertree building and refinement
    of the taxonomy without retention of additional information.
Consider the case, in which there are two phylogenetic inputs along with the taxonomy from
    tree Figure \ref{fig:Placing-one-incertae}b with the first phylogeny, $T_1$, providing the single
    rooted split: $V1\,X \mid \bullet\,U$ and the second phylogeny, $T_2$ providing the 
    rooted split: $W\, V1 \mid \bullet\,V2$.
If the supertree pipeline were run with just the taxonomy and and $T_1$ as inputs, then the supertree
    would be the tree shown in Figure \ref{fig:Placing-one-incertae}c.
One might hope that using that tree as the revised taxonomy and then adding the split from $T_2$ in
    a subsequent analysis would result in the same tree as if $T_1$ and $T_2$ to the original taxonomy.
While the topology for the final tree in either order of analyses would indeed be the tree shown in 
    Figure \ref{fig:Placing-one-incertae}b, the decision about whether or not to name $V$ would differ.
Our handling of \emph{incertae sedis} in the original taxonomy via rooted partial splits
    would encode the taxonomic definitions in the manner depicted in Figure \ref{fig:Placing-one-incertae-MRP}a,
    but the encoding of the revised taxonomy in Figure \ref{fig:Placing-one-incertae}c would result
    in an altered definition of taxon $V$, as shown in \ref{fig:Placing-one-incertae-MRP}c.
As expected, in the encoding of the revised taxonomy, the definition of $WX$ has been altered to reflect the
    fact $T_1$ placed $V$ inside $WX$.
Perhaps, unexpected is the fact that storing the revised taxonomy with an \emph{incertae sedis} annotation
    and then reprocessing the taxonomy using our rooted partial splits approach would also shift
    the definition of $V$ from a taxon that must exclude $W$ to one that may include $W$.

@@

@@

\newpage %TEMP

\begin{figure}
\subfloat[$\tau_{1}$]{\includegraphics[scale=0.5]{Figures/decomp/tree1}
}
\subfloat[$\tau_{2}$]{\includegraphics[scale=0.5]{Figures/decomp/tree2}
}
\subfloat[$\tau_{3}$]{\includegraphics[scale=0.5]{Figures/decomp/tree3}
}
\subfloat[$\tau_{4}$]{\includegraphics[scale=0.5]{Figures/decomp/tree4}
}

\subfloat[Taxonomy $\protect\taxonomy$]{\includegraphics[scale=0.5]{Figures/decomp/tax}
}
\subfloat[$Y$ and $Z$ decomposition]{\includegraphics[scale=0.5]{Figures/decomp/decompRep}
}
\subfloat[Pruned $Y$ and $Z$ decomposition]{\includegraphics[scale=0.5]{Figures/decomp/decompPruned}
}

\caption{\label{fig:decomp} Four phylogenetic trees (a-d) an
example taxonomy (e) which are possible inputs to the propinquity supertree pipeline.
(f) shows the taxonomic component of the subproblem decompositions for taxa $Y$ and $Z$ subproblems 
that would result
from an unmodified pipeline with used $\tau_1$ and either $\tau_2$ or $\tau_3$ as the phylogenetic
inputs.
(g) shows a possible decomposition from a decomposition tool that avoids duplication of taxa.
}
\end{figure}

@@

Simply allowing the subproblem decomposer to recognize that intrusion
    of an \emph{incertae sedis} taxon does not contest the existence
    of a clade would be straightforward, but problems arise if the
    decomposer is not further altered.
Consider, for example, the inputs of the $\tau_1$, $\tau_2$, $\tau_3$, and the taxonomy
    shown in Figure \ref{fig:decomp}.
Because the taxon $\alpha$ is annotated as being \emph{incertae sedis}, 
    neither  $\tau_1$, $\tau_2$, nor $\tau_3$ contest the monophyly of 
    the taxa $Y$, $Z$, or $\alpha$.
This would lead to the taxonomic portion of both the $Y$ and $Z$ subproblems
    containing the taxonomy for $\alpha$, leading to the members of
    $\alpha$ occurring in two different subproblems (see Figure \ref{fig:decomp}f).
Solving and merging these problems would result in the duplication of $\alpha$ 
    in the supertree.

One could easily
    imagine a ``pruning decomposer'' that does not allow the say taxon to
    occur in more than one subproblem.
This approach would work in a straightforward way if the phylogenetic inputs were only $\tau_1$ and $\tau_3$,
    yielding the decomposition shown in Figure \ref{fig:decomp}g, but if
    both $\tau_1$ and $\tau_2$ were inputs the decomposer would have to
    resolve conflict about the placement of $\alpha_1$.
Implementing the behavior in the subproblem decomposition step would
    break the separation between the ``divide'' and ``conquer'' steps
    which is a key aspect of propinquity's current efficiency.

Even considering ranked phylogenetic inputs of $\tau_4$ then $\tau_1$ from
    Figure \ref{fig:decomp} reveal some difficulties.
In this case the inclusion of $\alpha$ within the taxon $Y$ in $\tau_1$ would
    cause $\alpha$ to be solved in the subproblem for $Y$.
If $\tau_4$ were ranked higher than $\tau_1$, then the grouping of $Y_1+Y_2+\alpha$ would
    not be permitted because they conflict with the split $Y_1\,Y_3 \mid \bullet\,Y_2$ from
    $\tau_4$.
Thus, $\alpha$ would be placed inside of $Y$ on the basis of a rooted split
    that is not displayed in the final tree.
While this is placement is permissible based on the \emph{incertae sedis} annotation of the
    taxonomy, a major goal of the propinquity pipeline is to make it
    easy for biologists to understand the cause of different groupings in the 
    supertree.
Allowing splits which are not displayed to determine the placement of groups runs counter to that goal.

We choose to solve these problems by merging any subproblems that an
\emph{incertae sedis} taxon might be placed in.
The simplest way to
achieve this is simply to regard any taxon that has an \emph{incertae
sedis }taxon placed within it as contested.
This results in marking
both $B$ and $C$ as contested edges in the example above.
In fact,
this is the current behaviour of the non-\emph{incertae-sedis} aware
subproblem decomposer.
One downside of this approach is that, an \emph{incertae sedis} taxon
    place in only one input tree, the clade it is placed in will be
    marked as contested even though none of the counterintuitive
    behaviors mentioned above would arise if we did not
    mark it as contested.

@@
